\Title

### 2013-05-17 synmut.tex
Some aesthetic stuff from Richard which I had forgotten

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used by \citet{watts_architecture_2009} and 
thereby assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or were subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than those of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between  those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

@@

The above simulations show that hitchhiking can explain the observation of
deleterious mutations that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies -- $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous mutations at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
test two possible mechanisms that are biologically plausible and could explain
the transient rise of nonsynonymous mutations: time-dependent selection and
within-epitope competition.

@@

In the alternative hypothesis, several different escape mutations within the
same epitope might arise almost simultaneously and start to spread. Their
benefits are not additive, because each of them is essentially sufficient to
escape. As a consequence, several escape mutations rise to high frequency
rapidly, while the one with the smallest cost in terms of replication,
packaging, etc. is most likely to eventually fix. The emergence of multiple
sweeping nonsynonymous mutations in real HIV-1 infections has been shown
\citep{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
observed in  the evolution of resistance to 3TC, where the mutation 
M184V is often preceeded by M184I \citep{hedskog_dynamics_2010}. Similarly, AZT 
resistance often emerges via the competing TAM and TAM1 pathways.
Within epitope competition can be
implemented in the model through epistasis between escape mutations. While each
mutation is individually beneficial, combining the mutations is deleterious (no
extra benefit, but additional costs). Again, we find that the potential for
hitchhiking is little affected by within epitope competition but that the
fixation probability of nonsynonymous polymorphisms is reduced. With roughly six
mutations per epitope, the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios are not exclusive and possibly both
important in HIV-1 evolution.

>>>>

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases (the HIV-1 genome is a single stranded RNA)
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems.  We aligned the within-patient sequence samples 
to the reference NL4-3 strain used in \citet{watts_architecture_2009} and 
thereby assigned SHAPE reactivities to most positions in the alignment. 
We then calculated the distributions of SHAPE reactivities for synonymous 
polymorphisms that fixed or were subsequently lost (only polymorphisms with 
frequencies above 15\%).
As shown in \FIG{SHAPEA}, the reactivities of fixed alleles (red
histogram) are systematically larger than those of alleles that are lost (blue)
(Kolmogorov-Smirnov test on the cumulative distribution, $p\approx 0.002$). In
other words, alleles that are likely to break RNA helices are also more likely
to revert and finally be lost from the population. The average over all
mutations that are not observed (green) lies between  those that fix and
those that get lost. Note that this analysis will be sensitive only at positions
where the base pairing pattern of NL4-3 agrees with that of each patient's
initial consensus sequence (it is thus statistically conservative).

@@

The above simulations show that hitchhiking can explain the observation of
deleterious mutations that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous mutations at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous mutations suggests the rapid rise and fall of many alleles. We
test two possible mechanisms that are biologically plausible and could explain
the transient rise of nonsynonymous mutations: time-dependent selection and
within-epitope competition.

@@

In the alternative hypothesis, several different escape mutations within the
same epitope might arise almost simultaneously and start to spread. Their
benefits are not additive, because each of them is essentially sufficient to
escape and no additional benefit is gained from combining them.
As a consequence, several escape mutations rise to high frequency
rapidly, while the one with the smallest cost in terms of replication,
packaging, etc. is most likely to eventually fix, while all others are lost. The emergence of multiple
sweeping nonsynonymous mutations in real HIV-1 infections has been shown
\citep{moore_limited_2009, bar_early_2012}. This scenario has been explicitly 
observed in  the evolution of resistance to 3TC, where the mutation 
M184V is often preceeded by M184I \citep{hedskog_dynamics_2010}. Similarly, AZT 
resistance often emerges via the competing TAM and TAM1 pathways.
Within epitope competition can be
implemented in the model through epistasis between escape mutations. While each
mutation is individually beneficial, combining the mutations is deleterious (no
extra benefit, but additional costs). Again, we find that the potential for
hitchhiking is little affected by within epitope competition but that the
fixation probability of nonsynonymous polymorphisms is reduced. With roughly six
mutations per epitope, the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios are not exclusive and possibly both
important in HIV-1 evolution.
